Measurement Clusterization in D-optimal Designs for Bayesian Linear Inverse Problems

### 2023-05-07 latex/intro.tex
Add abstract (raw)

\section{Introduction}\label{section:intro}
Measurements play a fundamental role in generating observations and
are indispensable to any inference process. In the realm of data
science, selecting the optimal set of measurements holds particular
significance when inferring parameters of a physical process. Unlike
many other fields, where measurements are fixed or limited, in this
context, we have the freedom to choose which measurements to take. We
would like to take advantage of this freedom to choose measurements
that enhance accuracy, reduce costs, or achieve both objectives
simultaneously. Whether measurements involves specifying electrode
locations on the skin in electric impedance tomography
\cite{horesh2010impedance}, determining certain wavelengths in MRI
\cite{horesh2008mri}, or positioning sensors for detecting
ground-reflected waves in the search for oil
\cite{horesh2008borehole}, the selection of optimal measurements,
referred to as the problem of \emph{optimal design}, becomes
crucial. To achieve this, researchers optimize specific \emph{design
criteria}, with A- and D-optimality being two of the most widely
recognized and extensively studied design criteria
\cite{Chaloner1995}.

Interestingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  hooker2009, fedorov2012, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy inverse problem: inferring
the initial condition of the 1D heat equation (details in Section
\ref{section:example}). In Figure \ref{fig:clusterization
  illustration} D-optimal measurement locations are shown for
different numbers of allowed measurements. Notably, for six
measurement locations, the D-optimal design yields two sets of
measurements that are almost indistinguishable from one another. This
intriguing phenomenon is known as \emph{measurement clusterization}
\cite{Ucinski05}, and we refer to a design that exhibits measurement
clusterization as a \emph{clustered design}.

@@

    \end{axis}
  \end{tikzpicture}
  \caption{Measurement clusterization when inferring the initial
    condition of the 1D heat equation. Posterior pointwise standard
    deviations (lines) are plotted over the computational domain
    $\Omega = [0, \pi]$, for varying numbers of
    measurements. Measurement locations (circles) were chosen to
    according to the D-optimality criterion. Measurement
    clusterization occurs for six measurements: Only four measurement
    locations are visible, since two pairs of measurements are so
    close together they are indistinguishable.}
  \label{fig:clusterization illustration}
\end{figure}

Researchers widely agree that measurement clusterization is undesirable
\cite{fedorov1996, hooker2009, fedorov2012, Ucinski05,
  neitzel2019sparse}, prompting the exploration of various remedies to
address this issue. One approach involves merging ``close''
measurements \cite{fedorov2012}; however, this strategy merely
overlooks the phenomenon of measurement clusterization. An alternative
solution lies in \emph{clusterization-free design}s, where measurement
locations are deliberately chosen to be distant from one another. This
can be achieved by imposing distance constraints between measurements
or by introducing correlated errors that account for both observation
error and model misspecification \cite{Ucinski05}. For instance, in
the context of time-series analysis for pharmacokinetic experiments,
measurement clusterization can be mitigated by incorporating the modeling
of auto-correlation time within the noise terms
\cite{hooker2009}. %% This methodology shares a similar rationale with
%% the earlier approach, which assumed correlation between errors
%% \cite{Ucinski05}.

In spatial problems involving choice of measuremnts within a domain
$\Omega \subseteq \mathbb{R}^d, d=1,2,3$, many researchers circumvent
the problem of measurement clusterization by choosing measurements from a
coarse grid in $\Omega$ \cite{koval2020, alexanderian2021, attia2020,
  alexanderian2014, alexanderian2016,
  alexanderian2018efficient}. However, this approach incurs a
significant computational cost as it requires solving a difficult
combinatorial optimization problem to select optimal measurement
locations from a finite set. Typically, the optimization problem is
relaxed by first assigning optimal measurement "weights" in
$\mathbb{R}_+$ to potential measurement locations. Some authors
incorporate a sparsifying $\ell_1$ penalty term into the design
criterion, which is subsequently thresholded to achieve the desired
binary design \cite{horesh2008borehole}. Others progressively relax
the $\ell_1$ penalty to an $\ell_0$ penalty via a continuation method
\cite{alexanderian2016, alexanderian2014}. A clever alternative is to
cast the problem of finding optimal measurement weights as a stochastic
optimization problem \cite{attia2022stochastic}. All of the
aforementioned methods may indeed find a binary optimal design
restricted to a given coarse grid. However, none addresses one
fundamental issue: the restriction of measurement locations to a coarse
grid in $\Omega \subseteq \mathbb{R}^d, d=1,2,3$ inherently results in
a sub-optimal design.

Avoiding measurement clusterization is a pragmatic approach:
intuitively, we recognize that measurement clusterization is
undesirable, even though the underlying reasons may not be fully
clear. Consequently, we strive to prevent it and devise various
methodologies to avoid measurement clusterization. Yet each and every
one of these methodologies achieves this objective by imposing
restrictions on measurement locations, thereby fundamentally altering
the optimal design problem. Surprisingly, to the best of my knowledge,
no previous study has tried to address some seemingly simple yet
fundamental questions:
%
Why does measurement clusterization occur?
%
Why does imposing correlations between observations alleviate
measurement clusterization?
%
Should we aim to avoid measurement clusterization?
%
Can an optimal clustered design be relpaced with an equally optimal
non-clustered design?design?

\subsection{Contribution}
The primary objective of this study is to provide a comprehensive
understanding of measurement clusterization by addressing the
aforementioned questions. Our focus centers around investigating the
Bayesian D-optimality criterion, which involves maximizing the
expected Kullback-Leibler divergence between the posterior and prior
measures \cite{CoverThomas91, Chaloner1995}. To delve deeper into this
subject, we conduct an analysis of Bayesian D-optimal designs within
the context of linear inverse problems in Hilbert spaces. We propose a
novel relaxed model for D-optimality that maintains analytical
tractability and enables the identification of D-optimal designs using
Lagrange multipliers. This analytical framework facilitates the
exploration of the following questions posed at the end of the
previous paragraph:

@@

\item \label{q:replace} \textbf{Can an optimal clustered design be
  relpaced with an equally optimal non-clustered design?} In Section
  \ref{section:vanishing}, we answer this question in the affirmative,
  although we show that numerical experiments conducted using our
  model indicate a strong preference for clustered designs.
\end{enumerate}

The cornerstone of our investigation into measurement clusterization is
Theorem \ref{thm:char}, presented below and proven in Section
\ref{section:vanishing}. The key insight of Theorem \ref{thm:char}
lies in its fifth component, which highlights that a D-optimal design
aims to uniformly reduce posterior uncertainties across all posterior
covariance eigenvectors.

>>>>

\section{Introduction}\label{section:intro}
Measurements play a fundamental role in generating observations and
are indispensable to any inference process. In the realm of data
science, selecting the optimal set of measurements holds particular
significance when inferring parameters of a physical process. Unlike
many other fields, where measurements are fixed, in this context, we
have the freedom to choose which measurements to take. This freedom
should be taken advantage of to choose measurements that enhance
accuracy, reduce costs, or achieve both objectives
simultaneously. Whether measurements involve specifying electrode
locations on the skin in electric impedance tomography
\cite{horesh2010impedance}, determining certain wavelengths in MRI
\cite{horesh2008mri}, or positioning sensors for detecting
ground-reflected waves in the search for oil
\cite{horesh2008borehole}, the selection of optimal measurements,
referred to as the problem of \emph{optimal design}, becomes
crucial. To achieve this, researchers optimize specific \emph{design
criteria}, with A- and D-optimality being two of the most widely
recognized and extensively studied design criteria
\cite{Chaloner1995}.

Surprisingly, A- and D-optimal designs have been observed to yield
remarkably similar measurements in certain cases \cite{fedorov1996,
  hooker2009, fedorov2012, Ucinski05, neitzel2019sparse}. To
illustrate this point, we consider a toy inverse problem: inferring
the initial condition of the 1D heat equation (details in Section
\ref{section:example}). In Figure \ref{fig:clusterization
  illustration} D-optimal measurement locations are shown for
different numbers of measurements. Notably, for six measurements, a
D-optimal design yields two sets of measurements that are almost
indistinguishable from one another. This intriguing phenomenon is
known as \emph{measurement clusterization} \cite{Ucinski05}, and we
refer to a design that exhibits measurement clusterization as a
\emph{clustered design}.

@@

    \end{axis}
  \end{tikzpicture}
  \caption{Measurement clusterization for the inverse problem of
    inferring the initial condition of the 1D heat equation. Posterior
    pointwise standard deviations (lines) are plotted over the
    computational domain $\Omega = [0, \pi]$, for varying numbers of
    measurements. Measurement locations (circles) were chosen
    according to the D-optimality criterion. Measurement
    clusterization occurs for six measurements: Only four measurement
    locations are visible, since two pairs of measurements are so
    close together they are indistinguishable.}
  \label{fig:clusterization illustration}
\end{figure}

Researchers widely agree that measurement clusterization is
undesirable \cite{fedorov1996, hooker2009, fedorov2012, Ucinski05,
  neitzel2019sparse}, prompting the exploration of various remedies to
address this issue. One approach involves merging ``close''
measurements \cite{fedorov2012}; however, this strategy merely
overlooks the phenomenon of measurement clusterization. An alternative
solution lies in \emph{clusterization-free design}s, where measurement
locations are deliberately chosen to be distant from one another. This
can be achieved by imposing distance constraints between measurements
or by introducing correlated errors that account for both observation
error and model misspecification \cite{Ucinski05}. For instance, in
the context of time-series analysis for pharmacokinetic experiments,
measurement clusterization can be mitigated by incorporating the
modeling of auto-correlation time within the noise terms
\cite{hooker2009}.

In spatial problems involving choice of measuremnts within a domain
$\Omega \subseteq \mathbb{R}^d, d=1,2,3$, many researchers circumvent
the problem of measurement clusterization by choosing measurements
from a coarse grid in $\Omega$ \cite{koval2020, alexanderian2021,
  attia2020, alexanderian2014, alexanderian2016,
  alexanderian2018efficient}. However, this approach incurs a
significant computational cost as it requires solving a difficult
combinatorial optimization problem for choosing measurementst from a
finite set. The combinatorial optimization problem is usually relaxed
by first assigning optimal measurement "weights" in $\mathbb{R}_+$ to
the potential measurement locations. Some researchers incorporate a
sparsifying $\ell_1$ penalty term into the design criterion, which is
subsequently thresholded to achieve the desired binary design over the
coarse grid \cite{horesh2008borehole}. Others progressively relax the
$\ell_1$ penalty to an $\ell_0$ penalty via a continuation method
\cite{alexanderian2016, alexanderian2014}. A clever alternative is to
cast the problem of finding optimal measurement weights as a
stochastic optimization problem \cite{attia2022stochastic}. All of the
aforementioned methods may indeed find a binary optimal design
restricted to a given coarse grid. However, none addresses one
fundamental issue: the restriction of measurement locations to a
coarse grid in $\Omega \subseteq \mathbb{R}^d, d=1,2,3$ inherently
results in a sub-optimal design.

Avoiding measurement clusterization is a pragmatic approach:
intuitively, we recognize that measurement clusterization is
undesirable, even though the underlying reasons may not be fully
clear. Consequently, we strive to prevent it and devise various
methodologies to avoid measurement clusterization. Yet each and every
one of these methodologies achieves this objective by imposing
restrictions on measurement locations, thereby fundamentally altering
the optimal design problem. To the best of my knowledge, no previous
study has tried to address some seemingly simple yet fundamental
questions:
%
Why does measurement clusterization occur?
%
Why does imposing correlations between observations alleviate
measurement clusterization?
%
Should we aim to avoid measurement clusterization?
%
Is it possible to substitute an optimal clustered design with an
equally optimal non-clustered design?
%Can an optimal clustered design be relpaced with an equally optimal
%non-clustered design?

\subsection{Contribution}
The primary objective of this study is to provide a comprehensive
understanding of measurement clusterization by addressing the
aforementioned questions. Our focus centers around investigating the
Bayesian D-optimality criterion, which involves maximizing the
expected Kullback-Leibler divergence between the posterior and prior
measures \cite{CoverThomas91, Chaloner1995}. We conduct an analysis of
Bayesian D-optimal designs within the context of linear inverse
problems in Hilbert spaces. We propose a novel relaxed model for
D-optimality that maintains analytical tractability and enables the
identification of D-optimal designs using Lagrange multipliers. This
analytical framework facilitates the exploration of the questions
posed at the end of the previous paragraph:

@@

\item \label{q:replace} \textbf{Is it possible to substitute an
  optimal clustered design with an equally optimal non-clustered
  design?} In Section \ref{section:vanishing}, we answer this question
  in the affirmative, although we show that numerical experiments
  conducted using our model indicate a strong preference for clustered
  designs.
\end{enumerate}

The cornerstone of our investigation into measurement clusterization
is Theorem \ref{thm:char} proven in Section
\ref{section:vanishing}. The key insight of Theorem \ref{thm:char}
lies in its fifth component, which highlights that a D-optimal design
aims to uniformly reduce posterior uncertainties across all posterior
covariance eigenvectors. A similar conclusion was reached by Koval et
al.~\cite{koval2020}, who showed that A-optimal designs should be
constructed in the space of \emph{observations}.

### 2023-05-07 latex/paper.tex
Add abstract (raw)

\date{\today}
\begin{document}
\maketitle
\begin{abstract}
  We investigate the problem of measurement clusterization in D-optimal
  experimental design for infinite-dimensional Bayesian linear inverse
  problems. Many authors seem to believe that the fact that spatial
  correlations mitigateWe suggest an analytically tractable model for D-optimal
  designs. We argue that measurement clusterization in the case of iid
  observation noise is generic. We also prove that a model error term
  mitigates the clusterization phenomenon.

  %% As a part of the analysis we prove a Matrix Determinant
  %% Lemma in Hilbert spaces and a lemma for calculating
  %% $\frac{\der}{\der t} \log \det (I + X(t))$ for certain
  %% operator-valued functions. We also show how to decompose $M = AA^t$
  %% with $A$ constrained to have unit norm columns.
\end{abstract}

>>>>

\date{\today}
\begin{document}
\maketitle
\begin{abstract}
  \textbf{Setup} Taking measurements for inversion for physical
  processes requires expesive measurements.  \textbf{Problem}
  Unfortunately, when optimizing measurement locations, measurements
  get clustered.  \textbf{Why care} This a imortatn \textbf{What
    others do} Many researchers tried to avoind measuremnt
  clusterization, typically by modifying the optimization problem
  \textbf{What we do} We consider an analytically tractable model for
  Bayesian D-optimal design over Hilbert spaces. We prove that
  measurement clusterization poses no obstruction to D-optimality. We
  demonstrate that measuremnt clusterization is a generic property of
  D-optimal designs, and prove that measurement correlation mitigates
  the problem of measurement clusterization. We also give a full
  characterization of D-optimal designs, which, surprisingly, shows
  that D-optimal desings reduce uncertainty in the space of
  observations.  \textbf{Test on data} We test our conclusions via
  numeric implementation of our model
\end{abstract}

### 2023-05-07 latex/prelims.tex
Add abstract (raw)

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces (the subscripts p
and o are for ``parameter'' and ``observation'', respectively). Denote
$\hilo^* \simeq \hilo$ the Hilbert space of all linear functionals on
$\hilo$. Let $\fwd: \hilp \to \hilo$ the \emph{forward operator}. The
forward operator $\fwd$ is assumed linear and strongly smoothing (has
fast decaying modes --- the heat operator of the example in Section
\ref{section:intro} is a prime example). Take a Gaussian prior $\param
\sim \pr = \normal(\prmean ,\prcov)$ with some appropriate covariance
operator $\prcov$ on $\hilp$ \cite{Stuart10}. Note that $\fwd \prcov
\fwd^*$ is the prior covariance in $\hilo$ \cite{Stuart10}. As such,
$\fwd \prcov \fwd^*$ is assumed invertible --- an assumption which
will be used below. Let $m$ the number of measurements
taken. Measurements are taken via the \emph{measurement
operator}\footnote{It is common for the measurement and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires $\fwd$ and $\obs$ are explicitly separated
as in \cite{attia2022stochastic, cvetkovic2023choosing}.} $\obs \in (
\hilo^* )^m$, whose entries $\meas_j, j=1,\dots,m$ are called
\emph{measurement vector}s:

@@

The posterior measure $\post$ is Gaussian in this setting and its
covariance operator is:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

>>>>

\subsection{Bayesian Linear Inverse Problems}\label{subsec:abstract OED}
Let $\hilp$ and $\hilo$ be separable Hilbert spaces (the subscripts p
and o are for ``parameter'' and ``observation'', respectively). Denote
$\hilo^* \simeq \hilo$ the Hilbert space of all linear functionals on
$\hilo$. Let $\fwd: \hilp \to \hilo$ the \emph{forward operator}. The
forward operator $\fwd$ is assumed linear and strongly smoothing (has
fast decaying modes --- the heat operator of the example in Section
\ref{section:intro} is a prime example). Take a Gaussian prior $\param
\sim \pr = \normal(\prmean ,\prcov)$ with some appropriate covariance
operator $\prcov$ on $\hilp$ \cite{Stuart10}. Note that $\fwd \prcov
\fwd^*$ is the prior covariance in $\hilo$ \cite{Stuart10}. As such,
$\fwd \prcov \fwd^*$ is assumed invertible --- an assumption which
will be used below. Let $m$ the number of measurements
taken. Measurements are taken via the \emph{measurement
operator}\footnote{It is common for the measurement and forward
operators to be merged $\tmp := \obs \fwd$
\cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
following sections requires that $\fwd$ and $\obs$ are explicitly
separated as in \cite{attia2022stochastic, cvetkovic2023choosing}.}
$\obs \in ( \hilo^* )^m$, whose entries $\meas_j, j=1,\dots,m$ are
called \emph{measurement vector}s:

@@

It is useful to record that the posterior measure $\post$ is Gaussian
in this setting and its covariance operator is \cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

### 2023-05-07 latex/vanishing.tex
Add abstract (raw)

We are now ready to answer some of the questions posed in the
introduction:

@@

      \addlegendentry{prior}
      \addlegendentry{$o_1o_2$}
      \addlegendentry{$o_3o_4$}
      \addlegendentry{$o_5$}
      %% \addlegendentry{$o_4$}
      %% \addlegendentry{$o_5$}
      %% \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Clusterization (left, $\obs_1$ in \eqref{eq:two designs})
    and non-clusterization (right, $\obs_2$ in \eqref{eq:two designs})
    in D-optimal designs. Posterior precision per eigenvector is
    plotted for $\obs_1$ and $\obs_2$ from \eqref{eq:two
      designs}. Both designs are identical for all practical matters
    --- their posteriors are equal. In the left panel, a D-optimal
    design with clusterization is illustrated: $\meas_3 = \meas_5$. In
    the right panel, a D-optimal design without clusterization is
    illustrated. The contribution to the posterior precision due to
    $\meas_1$ and $\meas_2$ is denoted $\meas_1\meas_2$. These
    contributions are plotted together because $\meas_1$ and $\meas_2$
    are not in the direction of any eigenvector. Similarly for
    $\meas_3\meas_4$.}
  \label{fig:clusterization}
\end{figure}

>>>>

We are now ready to answer two of the questions posed in the
introduction:

@@

      \addlegendentry{prior}
      \addlegendentry{$o_1o_2$}
      \addlegendentry{$o_3o_4$}
      \addlegendentry{$o_5$}
      %% \addlegendentry{$o_4$}
      %% \addlegendentry{$o_5$}
      %% \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Clusterization (left, $\obs_1$ in \eqref{eq:two designs})
    and non-clusterization (right, $\obs_2$ in \eqref{eq:two designs})
    in D-optimal designs. Posterior precision per eigenvector is
    plotted for $\obs_1$ and $\obs_2$ from \eqref{eq:two
      designs}. Both designs are identical for all practical matters
    --- their posteriors are equal. In the left panel, a D-optimal
    design with clusterization is illustrated: $\meas_3 = \meas_5$. In
    the right panel, a D-optimal design without clusterization is
    illustrated. The contribution to the posterior precision due to
    $\meas_1$ and $\meas_2$ is denoted $\meas_1\meas_2$. These
    contributions are plotted together because neither $\meas_1$ nor
    $\meas_2$ are not in the direction of any eigenvector. Similarly
    for $\meas_3\meas_4$.}
  \label{fig:clusterization}
\end{figure}
